---
title: "lab_03"
author: "Derek Willis"
date: "2023-02-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
library(janitor)

# Load janitor
install.packages('janitor')

```

## Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564) |> clean_names()


# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv") |> clean_names()

payments

```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: 39 rows in pg_schools

```{r}
pg_schools <- payments |>
  filter(
    str_detect(vendor_name, 'PRINCE GEORGES PUBLIC') |
    str_detect(vendor_name, 'PG COUNTY PUBLIC') |
    str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL")
    )
pg_schools
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? What do you notice about the years listed, and what questions do you have about that data?

A2: FY 2020 had the highest total amount of payments to PGCPS.

```{r}
pg_schools |>
  group_by(fiscal_year) |>
  summarise(fiscal_total = sum(amount)) |>
  arrange(desc(fiscal_total))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have **only one row per vendor and zip code?** Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. Vendor names that are similar and should be combined are like "CMC REPRO & ASSOC INC" and "CMC REPRO & ASSOC., INC," "DIVERSIFIED PROPERTY SERVICES" and "DIVERSIFIED PROPERTY SERVICES INC", "A&I INC" and "A&I INC." and "FIRST POTOMAC ENVIRONMENTAL CORP INC" and "FIRST POTOMAC ENVIRONMENTAL CORPORATION."

To ensure that the data only has one row per vendor and zip code, it should be cleaned so that the different versions of some vendors' names are corrected and only counted as one name. To clean this data and make nearly identical vendors look identical, mutate() can be used to change the names of some of the vendors or merge two similar vendors into one. Using case_when() for one of these vendors only renames the vendor and keeps it as a separate row rather than creating one row that totals both of the amounts.

```{r}
payments |>
  distinct() |>
  filter(str_detect(agency_name, "CAPITAL PROJECT") & str_detect(agency_name, "UNIVERSITY OF MARYLAND")) |>
  group_by(vendor_name)|>
  summarise(total_amount = sum(amount))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: Stem cell research funding decreased since 2009 but has slowly started to increase in recent years, albeit while fluctuating.

```{r}
md_grants_loans 

md_grants_loans |>
  rename(source = grantor) |>
  mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description)) |> 
  filter(str_detect(description, "STEM CELL")) |>
  group_by(fiscal_year)|>
  summarise(total_amount = sum(amount))
```
