---
title: "Reverse Engineering Project Notebook"
output: html_notebook
---

Load libraries.

```{r}
options(scipen=999)

library(tidyverse)
library(readr)
library(janitor)
```

Load data. Change amount column from character to a number.

```{r}

#change Amount from <chr> to <dbl> by getting rid of special characters

sportsdonation <- read_csv("data/sports-political-donations.csv") |>
  mutate(Amount = str_replace_all(Amount,'\\$','')) |>
  mutate(Amount = str_replace_all(Amount, '\\,','')) |>
  mutate(Amount = as.numeric(Amount))
  

```

"American professional sports owners have contributed **nearly \$47 million in federal elections** since 2015, according to research by ESPN in partnership with FiveThirtyEight, including **\$10 million to Republican causes** and **\$1.9 million to Democratic causes** so far in the 2020 election cycle."

```{r}

#sum total donations: $46,978,697

sum(sportsdonation$Amount)

#find total Democrat donations 

sportsdonation |>
  filter(str_detect(Party, "Democrat")) 

#find total Republican donations 

sportsdonation |>
  filter(str_detect(Party, "Republican"))

```

"The research includes more than **160 owners** and commissioners spanning **125 teams**"

```{r}

#number of owners
sportsdonation |>
  group_by(Owner) |>
  summarise(n())



```

Who provided the highest number of donations? Charles Johnson.

```{r}

sportsdonation |>
  group_by(Owner) |>
  summarise(total_donations = n()) |>
  arrange(desc(total_donations))

```

Which league gave the most individual donations? MLB

```{r}

sportsdonation |>
  group_by(League) |>
  summarise(total_donation = n()) |>
  arrange(desc(total_donation))


```

How many donations for each party?

```{r}

sportsdonation |> 
  group_by(Party) |>
  summarise(n()) 

```
